---
title: "MCMC Simulation for Partial Order Inference"
author: "BayesianPO Package"
output:
  #pdf_document:
  #  toc: true
  html_document: default
  #  toc: true
  #  toc_float: true
  #  code_folding: show
  #  theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```



## (1) Setup and Installation

```{r load-packages}

# Load the BayesianPO package
source("../R/utilities.R")
source("../R/mcmc.R")
source("../R/mcmc_rj.R")  # Load reversible jump MCMC
source("../R/analysis.R")

library(MASS)
library(mvtnorm)
library(ggplot2)
library(igraph)

```

## (1) Data Generation

```{r}
X0 = rbind(c(1,2,3,4,5),c(1,3,2,4,5),c(1,2,4,3,5),c(1,2,4,3,5))
X = X0[sample(4,20,replace='T'),]
X[sample(20,5),4:5]=0

dada <- mx2list(X)
observed_orders = dada$observed_orders
choice_sets = dada$choice_sets
items <- dada$items
K <- ceiling(length(items)/2)
rng_seed = 1234
```

## (2) Standard MCMC Implementation

```{r fixed-mcmc}
# MCMC parameters
fixed_cfg <- list(
  iters            = 20000,                      # total MH iterations
  dr               = 0.10,                      # rho step
  rho_prior        = 1/6,                   # rho prior parameter
  noise_option     = "queue_jump",              # noise type
  noise_beta_prior = 9                        # queue-jump prob prior 
)

# 2.  Common arguments
args_fix <- list(
  observed_orders = observed_orders,
  choice_sets     = choice_sets,
  num_iterations  = fixed_cfg$iters,
  K               = K,
  dr              = fixed_cfg$dr,
  rho_prior       = fixed_cfg$rho_prior,
  noise_option    = "queue_jump",
  noise_beta_prior = fixed_cfg$noise_beta_prior,
  random_seed     = rng_seed
)

# 3.  Run the sampler
mcmc_fix <- do.call(mcmc_partial_order, args_fix)

```

## (6) Reversible Jump MCMC Implementation

The reversible jump MCMC allows the number of latent dimensions K to
vary during sampling, automatically determining the optimal model
complexity.

```{r rj-mcmc}
## ---- rj-mcmc -------------------------------------------------
# 1.  Configuration list ------------------------------------------------------
rj_cfg <- list(
  iters            = 20000, 
  dr               = 0.10,                      # rho step
  K_prior          = 3.0,                       # Poisson(λ) prior for K
  noise_option     = "queue_jump",              
  noise_beta_prior = 9,                         # used **only** for queue-jump
  rho_prior        = 1/6
)

args_common <- list(
  observed_orders = observed_orders,
  choice_sets     = choice_sets,
  num_iterations  = rj_cfg$iters,
  dr              = rj_cfg$dr,
  rho_prior       = rj_cfg$rho_prior,
  K_prior         = rj_cfg$K_prior,
  noise_option    = "queue_jump",
  noise_beta_prior = rj_cfg$noise_beta_prior,
  random_seed     = 42
)

# 2.  Run the sampler 
rj_results <- do.call(mcmc_partial_order_k, args_common)

```

## (7) Results and Diagnosics

```{r convergence, fig.width=12, fig.height=8}
# Plot log-likelihood traces
par(mfrow = c(1, 2))

# Check if we have valid data before plotting
plot(mcmc_fix$log_likelihood_trace, type = "l", 
       main = "Log-Likelihood (Fixed K)", xlab = "Iteration", ylab = "Log-Likelihood")

plot(rj_results$log_likelihood_trace, type = "l", 
       main = "Log-Likelihood (RJ-MCMC)", xlab = "Iteration", ylab = "Log-Likelihood")
```

We could select the burn-in period based on this.

```{r}
burn_in<-100
```

## (8) Parameter Estimation Comparison

```{r parameter-comparison}

# Fixed dimension estimates
rho_est_fixed <- mean(mcmc_fix$rho_trace[burn_in:length(mcmc_fix$rho_trace)])

# Reversible jump estimates  
rho_est_rj <- mean(rj_results$rho_trace[burn_in:length(rj_results$rho_trace)])
K_est_rj <- round(mean(rj_results$K_trace[burn_in:length(rj_results$K_trace)]))

# Create comparison table
comparison_df <- data.frame(
  Parameter = c("rho", "K"),
  Fixed_MCMC = c(rho_est_fixed,K),
  RJ_MCMC = c(rho_est_rj,  K_est_rj)
)

print(comparison_df)
```

## (9) Trace Plots

```{r trace-plots, fig.width=10, fig.height=8}
# Create comprehensive trace plots with prior/posterior comparisons

config <- list(
  prior = list(
    rho_prior = fixed_cfg$rho_prior,
    noise_beta_prior =  fixed_cfg$noise_beta_prior
  ),
  noise = list(
    noise_option = "queue_jump"
  )
)

plot_mcmc_results(mcmc_fix, 
                 config = config,
                 burn_in = 2)
```

## (10) RJ-MCMC Trace Plots

```{r rj-trace-plots, fig.width=12, fig.height=12}
# Create comprehensive trace plots for RJ-MCMC with K dimension
rj_config <- list(
  prior = list(
    rho_prior = rj_cfg$rho_prior,
    noise_beta_prior = rj_cfg$noise_beta_prior,
    K_prior = rj_cfg$K_prior
  ),
  noise = list(
    noise_option = "queue_jump"
  )
)

plot_mcmc_results(rj_results, 
                 config = rj_config,
                 burn_in = 2)
```


## (11) Dimension Selection Analysis

```{r dimension-analysis, fig.width=10, fig.height=6}
# Analyze K posterior distribution
K_posterior <- table(rj_results$K_trace[burn_in:length(rj_results$K_trace)])
K_probs <- K_posterior / sum(K_posterior)

# Plot K posterior
par(mfrow = c(1, 1))

barplot(K_probs, main = "Posterior Distribution of K", 
        xlab = "Number of Dimensions (K)", ylab = "Posterior Probability",
        col = "lightblue", border = "darkblue")
abline(v = K + 0.5, col = "red", lty = 2, lwd = 2)

```


# Model Comparison and Validation

## (1) Posterior Predictive Checks

```{r posterior-predictive}
# Calculate burn-in index (convert to trace index)

threshold <- 0.5

# Extract post-burn-in partial order traces
h_trace_fixed_post_burnin <- mcmc_fix$h_trace[burn_in:length(mcmc_fix$h_trace)]

h_trace_rj_post_burnin <- rj_results$h_trace[burn_in:length(rj_results$h_trace)]

# Compute posterior mean partial orders
cat("Computing posterior mean partial orders...\n")

# Fixed K MCMC: Average across all post-burn-in iterations
if (length(h_trace_fixed_post_burnin) > 0) {
  # Convert list of matrices to 3D array for easier averaging
  h_array_fixed <- array(0, dim = c(nrow(h_trace_fixed_post_burnin[[1]]), 
                                   ncol(h_trace_fixed_post_burnin[[1]]), 
                                   length(h_trace_fixed_post_burnin)))
  
  for (i in seq_along(h_trace_fixed_post_burnin)) {
    h_array_fixed[,,i] <- h_trace_fixed_post_burnin[[i]]
  }
  
  # Compute mean and apply threshold
  mean_h_fixed <- apply(h_array_fixed, c(1,2), mean)
  final_h_fixed <- (mean_h_fixed >= threshold) * 1
  
  # Apply transitive reduction to get minimal representation
  final_h_fixed <- transitive_reduction(final_h_fixed)
  
  cat("Fixed K MCMC: Averaged", length(h_trace_fixed_post_burnin), "post-burn-in samples\n")
} else {
  final_h_fixed <- matrix(0, nrow = nrow(h_true), ncol = ncol(h_true))
  cat("Warning: No post-burn-in samples for Fixed K MCMC\n")
}

# RJ-MCMC: Average across all post-burn-in iterations
if (length(h_trace_rj_post_burnin) > 0) {
  # Convert list of matrices to 3D array for easier averaging
  h_array_rj <- array(0, dim = c(nrow(h_trace_rj_post_burnin[[1]]), 
                                ncol(h_trace_rj_post_burnin[[1]]), 
                                length(h_trace_rj_post_burnin)))
  
  for (i in seq_along(h_trace_rj_post_burnin)) {
    h_array_rj[,,i] <- h_trace_rj_post_burnin[[i]]
  }
  
  # Compute mean and apply threshold
  mean_h_rj <- apply(h_array_rj, c(1,2), mean)
  final_h_rj <- (mean_h_rj >= threshold) * 1
  
  # Apply transitive reduction to get minimal representation
  final_h_rj <- transitive_reduction(final_h_rj)
  
  cat("RJ-MCMC: Averaged", length(h_trace_rj_post_burnin), "post-burn-in samples\n")
} else {
  final_h_rj <- matrix(0, nrow = nrow(h_true), ncol = ncol(h_true))
  cat("Warning: No post-burn-in samples for RJ-MCMC\n")
}

```

## (2) Visualize Estimated Partial Orders

```{r plot-estimates, fig.width=15, fig.height=5}
par(mfrow = c(1, 2))

plot_partial_order(
  final_h_fixed, items, title = "Fixed K MCMC Estimate\n(Posterior Mean)",
  vertex_size   = 38,
  edge_arrow_size = 0.6,
  label_cex     = 1.3,  # even bigger text
  frame_width   = 2.3   # bolder outlines
)


plot_partial_order(
  final_h_rj, items, title = "RJ-MCMC Estimate\n(Posterior Mean)",
  vertex_size   = 38,
  edge_arrow_size = 0.6,
  label_cex     = 1.3,  # even bigger text
  frame_width   = 2.3   # bolder outlines
)

```


## References

-   Nicholls, G. K., et al. (2024). Bayesian inference for partial
    orders. *arXiv preprint arXiv:2212.05524*.
-   Green, P. J. (1995). Reversible jump Markov chain Monte Carlo
    computation and Bayesian model determination. *Biometrika*, 82(4),
    711-732.

------------------------------------------------------------------------


